---
title: "Reproducible Research Course Project 1"
author: "Mlibazisi Ndlovu"
date: "19 July 2020"
output: html_document
---
Reproducible Research Project 1

The starting point is to load the fitness data

```{r setup, include=TRUE}
#create the working directory

if (!file.exists("data")) {
    dir.create("data")
}

# Fork/clone via git bash the following repsoitory which contains the data: https://github.com/rdpeng/RepData_PeerAssessment1

#read the data files into a data frame.
unzip(zipfile = "activity.zip")
activity = read.csv("activity.csv")
```

The next step is a quick overview of the data

```{r}

summary(activity)
dim(activity)
head(activity)
tail(activity)
```

Then clean and process the data 

```{r}
# Remove the rows with missing data

activity_clean <- na.omit(activity) 
```

```{r,include=FALSE}

# Load the tidyverse packages to be used for data procesing

library(tidyverse) 
```

```{r}
# create dataframes grouped by day and interval

activity_clean2 <- tbl_df(activity_clean)
act_date <- group_by(activity_clean2, date) 
act_interval <- group_by(activity_clean2, interval)
```

Calcualte the average total number of steps per day

```{r}
daily_steps <- summarise(act_date, sum(steps))
```

Plot a histogram of the number average total number of steps per day

```{r}
hist(daily_steps$`sum(steps)`,col = "blue",breaks = 9 ,main = "Total Steps Taken Per Day", xlab = "Number of steps", ylab = "Frequency")
```

Calculate the mean and median for the daily total number of steps

```{r}
mean(daily_steps$`sum(steps)`) 
median(daily_steps$`sum(steps)`)
```

Average daily activity pattern analysis

```{r}
#Obtain the average number of steps per interval acoss all the days
pattern <- summarise(act_interval, mean(steps))

# Plot the average total number of steps by interval
plot(pattern$interval,pattern$`mean(steps)`, type = "l",main = "Average Steps Taken Per Interval", xlab = "Interval", ylab = "Average Number of Steps" )

# Determine and report the interval with the most number of steps on average
w = as.character(pattern[which.max(pattern$`mean(steps)`),][1])
x = "The five minute interval containing the maximum steps on average is: "
print(c(x,w))
```

Perfoming analysis with the missing values being imputed

```{r}
 #Obtain and report the count of the missing values
Missing_data <- sum(is.na(activity))
Missing_data
y = "The number of missing values is: "
print(c(y,Missing_data))
```

```{r}
# Impute the missing values with the mean for each interval
avgsteps <- rep(pattern$`mean(steps)`,61) #create a vector of the interval averages 
activ_avgs <- cbind(activity,avgsteps) #create a new dataframe with a column for interval averages

activity_clean3 <- tbl_df(activ_avgs)
cleanactivity <- mutate(activity_clean3,steps_2 = ifelse(is.na(steps), avgsteps,steps)) #The new dataframe with missing values imputed in cloumn steps_2
```

Plot the histogram of daily total number of steps , wtih missing values imputed 

```{r}
act_date_2 <- group_by(cleanactivity, date) #group by date so as to analyse steps per day.

daily_steps_2 <- summarise(act_date_2, sum(steps_2))

hist(daily_steps_2$`sum(steps_2)`, col = "dark green",breaks = 9 ,main = "Total Steps Taken Per Day", xlab = "Number of steps", ylab = "Frequency")
```

Compare the means of the cleaned and uncleand data

```{r}
means <- c(mean(daily_steps$`sum(steps)`),mean(daily_steps_2$`sum(steps_2)`))
medians <- c(median(daily_steps$`sum(steps)`),median(daily_steps_2$`sum(steps_2)`))
comparison <- data.frame(means,medians, row.names = c("Raw Data"," Imputed Data"))
comparison
```

Activity patterns by day of the week

```{r}

#Add a column of days to the data

cleanactivity2 <- mutate(cleanactivity, Day =weekdays(as.Date(date)))
cleanactivity3 <- mutate(cleanactivity2, Daytype = ifelse(Day == "Saturday"|Day == "Sunday", "Weekend","Weekday"))
```

```{r}
#Group data by intervals

act_interval2 <- group_by(cleanactivity3, interval)
pattern2 <- summarise(act_interval2, mean(steps_2))
interval_avg <- rep(pattern2$`mean(steps_2)`,61)

cleanctivity4 <- mutate(cleanactivity3,interval_avg)
```

Plot a graph of activity over the weekdays and weekends

```{r}
g <- ggplot(cleanctivity4, aes(interval,interval_avg))
g + geom_line() + facet_grid(Daytype ~ .) + labs(title = "Activity Patterns By Time of Week") + labs(x ="Five Minute Interval", y = "Total Average Steps") + theme_light()
